---
title: "T-tests"
output: html_notebook
---

<link href="page_styles.css"></link>

# START with a flow chart
beep

```{r echo=F}
##note that you need RStudio 1.2 + to run the python code below

if(!require(reticulate)){install.packages("reticulate")}
##note that you (might?) need to install anaconda specifically so that you can call python (3.7)
library(reticulate)

## For notebook management
library(knitr) 

## NOTE - you need to use opts_knit for setting the directory if you are using a markdown or notebook. 
this.root.dir = "C:/Users/nt906822/OneDrive - University of Reading/Github/psychOpenStatsbook/"
opts_knit$set(root.dir = this.root.dir)

##if not using a markdown or notebook, use the following
setwd("C:/Users/nt906822/OneDrive - University of Reading/Github/psychOpenStatsbook/")


```


T-tests are useful when we want to make comparisons between two conditions. These can either be <b>between
groups</b> who are in different conditions, or <b>within groups</b> for participants who experience both conditions. 

Let's work on t-tests for Cats and Dogs first. For this section, we are going to imagine that we are investigating whether cats or dogs are <b>taller</b>. This is a <b>between-subjects</b> or <b>independent samples</b> t-test. Here's the raw data (in cm):

```{r echo=F}
cat_dog_data <- read.csv("cats_dogs_height.csv")
cat_dog_data
```

Before we do anything else, let's check whether the dogs are taller than the cats (which is what we'd expect). Let's get the means and standard deviations for each animal.

-----
<div id="means_div">
  <button class="btn btn-primary hide_show">SPSS</button>
  <button class="btn btn-primary hide_show">R</button>
  <button class="btn btn-primary hide_show">Python</button>
  <div class="spss_div">
## Mean using SPSS
  </div>
  <div class="r_div">
## Mean using R
  
```{r}
## let's load the data and separate it

cat_dog_data <- read.csv("cats_dogs_height.csv")
cat_data = cat_dog_data$height[cat_dog_data$animal == "cat"]
dog_data = cat_dog_data$height[cat_dog_data$animal == "dog"]
```

You'll see above a couple of steps to prepare the data for us getting the mean and standard deviations for the cat and dog data separately. Let's get the means for each:

```{r}
mean(cat_data)
```

```{r}
mean(dog_data)
```

  </div>
  <div class="python_div">
## Mean using python

First, let's prepare the data by loading the relevant packages and storing the data into a dataframe using the <b>pandas</b> package:

```{python}
import os
os.chdir("C:/Users/nt906822/OneDrive - University of Reading/Github/psychOpenStatsbook/")
import statistics   # for it's function "mean"
import pandas as pd # to read the csv into a dataframe

cat_dog_data = pd.read_csv('cats_dogs_height.csv')
cat_data = cat_dog_data["height"][cat_dog_data["animal"] == "cat"]
dog_data = cat_dog_data["height"][cat_dog_data["animal"] == "dog"]

```

Now we can focus on getting the mean height for the cats and dogs. 

Let's start with cats

```{python}
statistics.mean(cat_data)
```
```{python}
#don't run the above one or else the html file will not show the code
statistics.mean(cat_data)
```
And then dogs
```{python}
statistics.mean(dog_data)
```
```{python}
#don't run the above one or else the html file will not show the code
statistics.mean(dog_data)
```
  </div>
</div>

-----

So we now know that dogs were taller than cats (unsurprisingly). Let's use t-tests to investigate whether this difference is statistically significant.


<div id="ind_samples_div">
  <button class="btn btn-primary hide_show">SPSS</button>
  <button class="btn btn-primary hide_show">R</button>
  <button class="btn btn-primary hide_show">Python</button>
  <div class="spss_div">
## Independent samples t-test using SPSS
  </div>

  <div class="r_div">
## Independent samples t-test using R

```{r}

cat_dog_data <- read.csv("cats_dogs_height.csv") # in case you haven't already loaded the data

cat.data = cat_dog_data$height[cat_dog_data$animal == "cat"]
dog.data = cat_dog_data$height[cat_dog_data$animal == "dog"]

t.test(cat.data,      #group 1s data
       dog.data,      #group 2s data
       paired = F,    #confirming that this is between-subjects/independent samples
       var.equal = T) #we are currently assuming that the variance is equal between groups, and so can run a student's t-test. HOWEVER, we will challenge this assumption in a moment.

```
  As we can see in the output above, there's a VERY significant difference in heights between cats and dogs. Let's extract the most helpful snippets of information by storing this t-test as a variable and recalling the key details.
  
```{r}
cat_dog_t_test <- t.test(cat.data,      #group 1s data
                         dog.data,      #group 2s data
                         paired = F,    #confirming that this is between-subjects/independent samples
                         var.equal = T) #we are currently assuming that the variance is equal between groups, and so can run a student's t-test. HOWEVER, we will challenge this assumption in a moment.

```
Let's start with the p-value
```{r}
cat_dog_t_test$p.value
```
The p-value is much less than the .05 threshold needed for a significant result. In fact, we would write this as p < .001.

Let's move on to the t-value:

```{r}
cat_dog_t_test$statistic
```

This is a high t-value, and interestingly it's negative. All this means is that the second condition had a higher mean than the first condition.
  
  
  </div>
  <div class="python_div">
## Independent samples t-test using Python

```{python}
#print(a)
##note that print() function blocks presenting of the python code!!

a=5
```
  </div>
</div>



```{r}
sd(cat_data)
```
```{r}
sd(dog_data)
```
  

# End of chapter questions!

# would be good to have a way that there are MANY questions that users can contribute to help students consolidate their understanding. May need an elegant way to avoid this clogging up the ebooks though...


<script src="page_management.js"></script>
